Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-07-07 survey/survey.tex
finished with final pass

\documentclass{article}
\usepackage[utf8x]{inputenc}
\usepackage{agda}
\usepackage[nocenter]{qtree}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{listings}
\usepackage{multicol}
\usepackage{stmaryrd}
\usepackage{proof}
\usetikzlibrary{quotes,fit,positioning}
\tikzset{>=latex}

@@

\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
%% \newcommand{\lid}{\textsf{lid}}
\newcommand{\alt}{~|~}
%% \newcommand{\rid}{\textsf{rid}}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\ensuremath{\mathit{unite}_+\mathit{l}}}
\newcommand{\identrp}{\ensuremath{\mathit{uniti}_+\mathit{l}}}
\newcommand{\identlsp}{\ensuremath{\mathit{unite}_+\mathit{r}}}
\newcommand{\identrsp}{\ensuremath{\mathit{uniti}_+\mathit{r}}}
\newcommand{\swapp}{\ensuremath{\mathit{swap}_+}}
\newcommand{\assoclp}{\ensuremath{\mathit{assocl}_+}}
\newcommand{\assocrp}{\ensuremath{\mathit{assocr}_+}}
\newcommand{\identlt}{\ensuremath{\mathit{unite}_*\mathit{l}}}
\newcommand{\identrt}{\ensuremath{\mathit{uniti}_*\mathit{l}}}
\newcommand{\identlst}{\ensuremath{\mathit{unite}_*\mathit{r}}}
\newcommand{\identrst}{\ensuremath{\mathit{uniti}_*\mathit{r}}}
\newcommand{\swapt}{\ensuremath{\mathit{swap}_*}}
\newcommand{\assoclt}{\ensuremath{\mathit{assocl}_*}}
\newcommand{\assocrt}{\ensuremath{\mathit{assocr}_*}}
\newcommand{\absorbr}{\ensuremath{\mathit{absorbr}}}
\newcommand{\absorbl}{\ensuremath{\mathit{absorbl}}}
\newcommand{\factorzr}{\ensuremath{\mathit{factorzr}}}
\newcommand{\factorzl}{\ensuremath{\mathit{factorzl}}}
\newcommand{\dist}{\ensuremath{\mathit{dist}}}
\newcommand{\factor}{\ensuremath{\mathit{factor}}}
\newcommand{\distl}{\ensuremath{\mathit{distl}}}
\newcommand{\factorl}{\ensuremath{\mathit{factorl}}}
%% \newcommand{\distz}{\mathit{absorbr}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\sem}[1]{\ensuremath{\llbracket{#1}\rrbracket}}

@@

%%  Something about Theseus?
\end{abstract}

% * Reversibility intro / motivation as we  have now more or less

% * Thesis statement: programs are reversible deformations on data /
%    spaces, almost necessarily by definition

% * Focus now is on data

% * If data is plain finite sets, programs become permutations (deformations on finite sets).

% * If data is structured trees, we get Pi.

% * Explain Pi with examples.

% * If data is itself deformations-on-finite-sets, then the
%   deformations between them become something quite
%   interesting. Explain Pi level 2 with examples.

% * Conclude with thoughts regarding other kinds of data that can be
%   plugged in into that story.

@@

% \jc{The notation here is somewhat inconsistent with the above,
% mainly in the notation for values.}

@@

%% Random idea:
%% for a product monoidal this is wave-style
%% for a sum monoidal this is particle-style
%% when we have both we do have a wave-particle duality

@@

\begin{comment}
\jc{An additional ``problem'' is that since our circuits have a
left-to-right orientation, $A * B$ drawn as boxes left-to-right
gives an implicit ordering that is going to be hard to ignore.
I'm starting to think that the diagrams in the MO answer might
be better: use a colour (or like in Selinger's paper, at the
very end), symbols in the nodes that split.  So both would
be displayed as vertical stacks (only). Because I think that
swap* is too hard to justify in this notation. Of course,
if we could do things in 3D and $*$ was displayed using
(say) depth, that would be different again.}

This idea leads to a resource-based interpretation of types as
\emph{possible events} and of the values as \emph{consistent timelines
  of events} as explained below. The event $1$ is trivial, and the
event $0$ is impossible.  We visualize the denotation of $1$ as a
point in space, and $0$ as empty space. The denotation of $A + B$ is
the denotation of~$A$ drawn on top of the denotation of $B$ and
separated by a border, and the denotation of $A * B$ is just the
denotation of $A$ written next to the denotation of $B$.  For example,
let $2$ be the type $1 + 1$ represented by a stack of two boxes each
with a point inside it. The type $(2 + 2) * (2 + 2)$ would then be two
such stacks laid next to each other. Composing further, the type
$((2 + 2) * (2 + 2)) + ((2 + 2) * (2 + 2))$ is built by stacking the
last two diagrams as shown below on the left:
\begin{center}
\hfill
\includegraphics[scale=0.3]{images/tall-tl-1}
~
\includegraphics[scale=0.3]{images/tall-tl-2}
\hfill (*)
\end{center}
If we call the type on the left $A$ then the full diagram above
represents the type $A*A$.

To understand how diagrams (types) can transform through the
evaluation of  a program, we need to reason about the equivalence of
different diagrams. Consider for example, the two diagrams below:

\begin{center}
\hfill
\includegraphics[scale=0.3]{images/distrib-1}
\qquad \qquad \qquad \qquad \qquad
\includegraphics[scale=0.3]{images/distrib-2}
\hfill (**)
\end{center}

The diagram on the left represents the type $A * (B + C)$ while the
one on the right represents the type $(A * B) + (A * C)$. These two
types are equivalent in $\Pi$ but why are they equivalent in the
graphical notation? We argue that they admit the same consistent
timelines where a consistent timeline is defined recursively on the
structure of a diagram:
\begin{itemize}
\item An endpoint (the denotation of $1$) has exactly one timeline
  through it.  Empty space (the denotation of $0$) has no timelines
  through it.
\item A timeline through a stack of alternatives
  $A_1 + A_2 + \cdots + A_n$ is a timeline through exactly one of the
  possible alternatives.
\item A timeline through a spatial composition of cells
  $A_1 * A_2 * \cdots * A_n$ is a timeline through \emph{all} of the
  cells, in some order.
\end{itemize}

For example, the red line in the diagram (*) represents a consistent
timeline. For the diagrams (**), we note that any consistent timeline
for the left diagram must go through $A$ and either one of $B$ or $C$
and for the diagram on the right, the choice happens earlier but the
result is the same: any consistent timeline must go through either $A$
and $B$ or through $A$ and $C$.

Other than distributivity, $\Pi$ combinators simply allow us to shuffle
the components of a stack around as well as shuffle the positions of
components in the plane.
\end{comment}
% \amr{
%   Let’s have a geometric representation of spaces:
% - A + B means putting A next to B
% - A * B means putting A and B on two perpendicular axes and filling
% the square (volume) between them
% All the combinators would indeed have a natural action and if we have
% a few nice pictures it would be great. What do you think? }

% \newcommand{\graphzero}{}
% \newcommand{\graphone}{\ensuremath{\bullet}}
% \newcommand{\graphplus}[2]{\node (top) {#1}; \node [below of=top] {#2};}

% \begin{center}
% \begin{tikzpicture} %% [declare function = {plus(\x,\y) = \node (top){\x}; \node [below of=top] {\y};}]
% \node (x) {\graphone};
% \node (y) {\graphone};
% %% {plus(\x,\y)}
% \end{tikzpicture}
% \end{center}

% \begin{definition}{Primitive operators of ~\ensuremath{\Pi } }
% \label{def:primitives-langRev}

% \[\begin{array}{rrcll}
%  \identlp :& 0 + b &\leftrightarrow & b &: \identrp \\
%  \swapp :& b_1 + b_2 &\leftrightarrow & b_2 + b_1 &: \swapp \\
%  \assoclp :& b_1 + (b_2 + b_3) &\leftrightarrow & (b_1 + b_2) + b_3 &: \assocrp \\
%  \identlt :& 1 \times b &\leftrightarrow & b &: \identrt \\
%  \swapt :& b_1 \times b_2 &\leftrightarrow & b_2 \times b_1 &: \swapt \\
%  \assoclt :& b_1 \times (b_2 \times b_3) &\leftrightarrow & (b_1 \times b_2) \times b_3 &: \assocrt \\
%  \absorbr :& 0 \times b &\leftrightarrow & 0 &: \factorzl \\
%  \dist :& (b_1 + b_2) \times b_3 &\leftrightarrow & (b_1 \times b_3) + (b_2 \times b_3) &: \factor \\
%  \end{array}\]
% %subcode source isomorphisms.tex:427

% \end{definition}

% Each line of this table is to be read as the definition of one or two
% operators. For example, corresponding to the \textit{identity of
%   \ensuremath{\times}} isomorphism (\ensuremath{1\times b\leftrightarrow b}) we have the two operators
% \ensuremath{\identlt:1\times b\leftrightarrow b} (reading the isomorphism from left to right) and
% \ensuremath{\identrt:b\leftrightarrow 1\times b} (reading the isomorphism from right to
% left). These operators are inverses of each other.  Each of the two
% cases of commutativity defines one operator that is its own inverse.

% Having defined primitive operators, we need some means of composing
% them. We construct the composition combinators out of the closure
% conditions for isomorphisms.

% \begin{definition}{Composition in \ensuremath{\Pi } }
% $$
% \infer{ \idc : b \leftrightarrow b}{
% 	 ~
% }
% \quad
% \infer{ \ !~c : b_2 \leftrightarrow b_1}{
% 	 c : b_1 \leftrightarrow b_2
% }
% \quad
% \infer{ c_1\odot c_2 : b_1 \leftrightarrow b_3}{
% 	 c_1 : b_1 \leftrightarrow b_2
% 	&
% 	 c_2 : b_2 \leftrightarrow b_3
% }
% $$
% $$
% \infer{ c_1 + c_2 : b_1 + b_2 \leftrightarrow b_3 + b_4}{
% 	 c_1 : b_1 \leftrightarrow b_3
% 	&
% 	 c_2 : b_2 \leftrightarrow b_4
% }
% \quad
% \infer{ c_1 \times c_2 : b_1 \times b_2 \leftrightarrow b_3 \times b_4}{
% 	 c_1 : b_1 \leftrightarrow b_3
% 	&
% 	 c_2 : b_2 \leftrightarrow b_4
% }
% $$
% %subcode source isomorphisms.tex:462
% \end{definition}

% Thus we have program constructs that witness reflexivity \ensuremath{\idc},
% symmetry \ensuremath{\mathit{sym}}, and transitivity~\ensuremath{\odot} and two parallel
% composition combinators, one for sums~\ensuremath{+} and one for
% pairs~\ensuremath{\times}.

@@

\begin{center}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
%%  \node at (3.3,-2) {$\mathit{swapl}_{+\Leftrightarrow}$} ;
%%  \node at (2.5,-1.3) {$((c_2~\oplus~c_1)~\odot~\mathit{swap}_{+})$};
%%  \node at (2.5,-2.7) {$(\mathit{swap}_{+}~\odot~(c_1~\oplus~c_2))$};
%  \draw (-2,-2) ellipse (0.5cm and 1cm);
  \draw[fill] (-2,-1.5) circle [radius=0.025];
  \node[below] at (-2.1,-1.5) {$A$};
  \node[below] at (-2.1,-1.9) {$+$};
  \draw[fill] (-2,-2.5) circle [radius=0.025];
  \node[below] at (-2.1,-2.5) {$B$};

%  \draw (6.5,-2) ellipse (0.5cm and 1cm);
  \draw[fill] (6.5,-1.5) circle [radius=0.025];
  \node[below] at (6.7,-1.5) {$C$};
  \node[below] at (6.7,-1.9) {$+$};
  \draw[fill] (6.5,-2.5) circle [radius=0.025];
  \node[below] at (6.7,-2.5) {$D$};

@@

  \path (1.5,0.5) node (tc1) [func] {$c_1$};
  %\draw     (1,0.9)  -- (2,0.9) -- (2,0.1) -- (1,0.1) -- cycle ;
  %\node at (1.5,0.5) {$c_1$};

  \path (1.5,-0.5) node (tc2) [func] {$c_2$};
  %\draw     (1,-0.1)  -- (2,-0.1) -- (2,-0.9) -- (1,-0.9) -- cycle ;
  %\node at (1.5,-0.5) {$c_2$};

  \path (3,-4.5) node (bc1) [func] {$c_1$};
  %\draw     (2.5,-4.1)  -- (3.5,-4.1) -- (3.5,-4.9) -- (2.5,-4.9) -- cycle ;
  %\node at (3,-4.5) {$c_1$};

  \path (3,-3.5) node (bc2) [func] {$c_2$};
  %\draw     (2.5,-3.1)  -- (3.5,-3.1) -- (3.5,-3.9) -- (2.5,-3.9) -- cycle ;
  % \node at (3,-3.5) {$c_2$};

\end{tikzpicture}
\end{center}
The top path is the $\Pi$ program
$(c_1~\oplus~c_2)~\odot~\swapp$ which deforms the
type $A$ by $c_1$, deforms the type $B$ by $c_2$, and deforms the
resulting space by a twist that exchanges the two injections into the
sum type. The bottom path performs the twist first and then deforms
the type $A$ by $c_1$ and the type $B$ by $c_2$ as before. One
could imagine the paths are physical \emph{elastic} wires in $3$ space, where
the deformations $c_1$ and $c_2$ as arbitrary deformations on these wires, and
the twists do not touch but are in fact well-separated. Then, holding the
points $A$, $B$, $C$, and $D$ fixed, it is possible to imagine
sliding $c_1$ and $c_2$ from the top wire rightward past the
twist, and then using the elasticity of the wires, pull the
twist back to line up with that of the bottom --- thus making
both parts of the diagram identical.  Each of these moves
can be undone (reversed), and doing so would take the bottom
part of the diagram into the top part.  In other
words, there exists a deformation of the program
$(c_1~\oplus~c_2)~\odot~\swapp$ to the program
$\swapp \odot (c_2~\oplus~c_1)$. We can also show that this
means that, as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and
$\swapp \odot (c_2~\oplus~c_1)$ are equal. And, of course, not
all programs between the same types can be deformed into one
another. The simplest example of inequivalent deformations
are the two automorphisms of $1+1$, namely $\idc$ and $\swapp$.

@@

Previously we saw how we could take the proof terms of commutative semirings
equivalences as our starting point for $\Pi$. What we need
now is to understand how \emph{proofs} of algebraic identities should be
considered equivalent. Classical algebra does not help, as proofs
are not considered first-class citizens. However,
another route is available to us: since the work of
Hofmann and Streicher~\cite{hofmann96thegroupoid}, we know that
one can model types as \emph{groupoids}.  The additional
structure comes from explicitly modeling the ``identity
types'': instead of regarding all terms which witness
the equality of (say) $a$ and $b$ of type $A$ as being
indistinguishable, we posit that there may in fact be many.
This consequences of this one decision are enough to show that
types can be modeled by groupoids.

Thus, rather than looking at (untyped) commutative semirings,
we should look at a \emph{typed} version. This process frequently
goes by the moniker of ``categorification.''  We want a categorical
algebra, where the basic objects are groupoids (to model our types),
and where there is a natural notion of $+$ and $*$.  At first,
we hit what seems like a serious stumbling block: the category of
all groupoids, \Gpd, does not have either co-products
nor products. However, we don't want to work internally in
\Gpd -- we want operations
\emph{on} groupoids. In other words, we want something akin to
symmetric monoidal categories, but with two interacting
monoidal structures.  Luckily, this already exists: the categorical
analog to (commutative) semirings are (symmetric) Rig
Categories~\cite{laplaza72,kelly74}.
This straightforwardly generalizes to symmetric Rig Groupoids.

How does this help? Coherence conditions! Symmetric monoidal categories,
to start somewhere simple, do not just introduce natural transformations
like the associator $\alpha$ and the left and right unitors ($\lambda$
and $\rho$ respectively), but also coherence conditions that these must satisfy.
Looking, for example, at just the additive fragment of $\Pi$ (i.e. with just $0$,
$1$ and $+$ for the types, $\odot$ and $\oplus$ as combinators, and
only the terms so expressible), the sub-language would correspond,
denotationally, to exactly (non-empty) symmetric monoidal groupoids. And what
these possess are exactly some \emph{equations between equations}
as commutative diagrams.  Transporting these coherence conditions, for
example those that express that various transformations are \emph{natural}
to $\Pi$, gives a list of equations between $\Pi$ programs.
Furthermore, all the natural transformations
that arise are in fact natural \emph{isomorphisms} -- and thus
reversible.

@@

Starting with the simplest constructions, the first two constructs in
Fig.~\ref{fige} are the level-2 analogs of $+$ and $*$, which respectively
model level-1 choice composition and parallel composition (of
equivalences).  These allow us to ``build up'' larger equivalences from smaller
ones.  The next two express that both of these composition operators distribute
over sequential composition $\odot$ (and vice versa).

@@

The bottom line in Fig.~\ref{figj} is actually a linear
restatement of the famous ``pentagon diagram'' stating a
particular coherence condition for monoidal categories~\cite{KELLY197197}.
To make the relation between $\Pi$ as a language and the
language of category theory, the figure below
the same morphism but in categorical terms.

@@

\begin{comment}
For example, the last equivalence captures the idea that the
program $\assocrt \odot \assocrt$ transforms an input of
type $((t_1 * t_2) * t_3) * t_4$ to an output of type
$t_1 * (t_2 * (t_3 * t_4))$ as follows:

\begin{center}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]
\node[above] (A) at (-2,0) {$t_1$};
\node[above] (B) at (-1.5,0) {$t_2$};
\node[above] (C) at (-1,0) {$t_3$};
\node[above] (D) at (-0.5,0) {$t_4$};
\draw[thick] (A) -- (-2,-1.5);
\draw[thick] (B) -- (-1.5,-1.5);
\draw[thick] (C) -- (-1,-1.5);;
\draw[thick] (D) -- (-0.5,-1.5);
\draw[red,densely dashed] (-1.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
\draw[green,densely dashed] (-1.6,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
\end{tikzpicture}
$\leadsto$
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]
\node[above] (A) at (-2,0) {$t_1$};
\node[above] (B) at (-1.5,0) {$t_2$};
\node[above] (C) at (-1,0) {$t_3$};
\node[above] (D) at (-0.5,0) {$t_4$};
\draw[thick] (A) -- (-2,-1.5);
\draw[thick] (B) -- (-1.5,-1.5);
\draw[thick] (C) -- (-1,-1.5);;
\draw[thick] (D) -- (-0.5,-1.5);
\draw[red,densely dashed] (-1.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
\draw[green,densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
\end{tikzpicture}
$\leadsto$
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]
\node[above] (A) at (-2,0) {$t_1$};
\node[above] (B) at (-1.5,0) {$t_2$};
\node[above] (C) at (-1,0) {$t_3$};
\node[above] (D) at (-0.5,0) {$t_4$};
\draw[thick] (A) -- (-2,-1.5);
\draw[thick] (B) -- (-1.5,-1.5);
\draw[thick] (C) -- (-1,-1.5);;
\draw[thick] (D) -- (-0.5,-1.5);
\draw[red,densely dashed] (-0.9,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
\draw[green,densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
\end{tikzpicture}
\end{center}
% \[
% ((t_1 * t_2) * t_3) * t_4 \leadsto
% (t_1 * t_2) * (t_3 * t_4) \leadsto
% t_1 * (t_2 * (t_3 * t_4))
% \]

\noindent The program on the right
$((\assocrt \otimes \idc) \odot \assocrt) \odot (\idc \otimes
\assocrt)$ achieves the same end-to-end transformation using the
following sequence of steps:

\begin{center}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]]
\node[above] (A) at (-2,0) {$t_1$};
\node[above] (B) at (-1.5,0) {$t_2$};
\node[above] (C) at (-1,0) {$t_3$};
\node[above] (D) at (-0.5,0) {$t_4$};
\draw[thick] (A) -- (-2,-1.5);
\draw[thick] (B) -- (-1.5,-1.5);
\draw[thick] (C) -- (-1,-1.5);;
\draw[thick] (D) -- (-0.5,-1.5);
\draw[red,densely dashed] (-1.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
\draw[green,densely dashed] (-1.6,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
\end{tikzpicture}
$\leadsto$
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]]
\node[above] (A) at (-2,0) {$t_1$};
\node[above] (B) at (-1.5,0) {$t_2$};
\node[above] (C) at (-1,0) {$t_3$};
\node[above] (D) at (-0.5,0) {$t_4$};
\draw[thick] (A) -- (-2,-1.5);
\draw[thick] (B) -- (-1.5,-1.5);
\draw[thick] (C) -- (-1,-1.5);;
\draw[thick] (D) -- (-0.5,-1.5);
\draw[red,densely dashed] (-1.25,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
\draw[green,densely dashed] (-1.5,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
\end{tikzpicture}
$\leadsto$
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]]
\node[above] (A) at (-2,0) {$t_1$};
\node[above] (B) at (-1.5,0) {$t_2$};
\node[above] (C) at (-1,0) {$t_3$};
\node[above] (D) at (-0.5,0) {$t_4$};
\draw[thick] (A) -- (-2,-1.5);
\draw[thick] (B) -- (-1.5,-1.5);
\draw[thick] (C) -- (-1,-1.5);;
\draw[thick] (D) -- (-0.5,-1.5);
\draw[red,densely dashed] (-1.25,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
\draw[green,densely dashed] (-1.0,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
\end{tikzpicture}
$\leadsto$
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]]
\node[above] (A) at (-2,0) {$t_1$};
\node[above] (B) at (-1.5,0) {$t_2$};
\node[above] (C) at (-1,0) {$t_3$};
\node[above] (D) at (-0.5,0) {$t_4$};
\draw[thick] (A) -- (-2,-1.5);
\draw[thick] (B) -- (-1.5,-1.5);
\draw[thick] (C) -- (-1,-1.5);;
\draw[thick] (D) -- (-0.5,-1.5);
\draw[green,densely dashed] (-0.9,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
\draw[red,densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
\end{tikzpicture}
\end{center}
% \[
% ((t_1 * t_2) * t_3) * t_4 \leadsto
% (t_1 * (t_2 * t_3)) * t_4 \leadsto
% t_1 * ((t_2 * t_3) * t_4) \leadsto
% t_1 * (t_2 * (t_3 * t_4))
% \]

% \jc{I am not a big fan of these pictures -- mainly because I can't give any
% intuition based on them of why these can be smoothly deformed into one another.
% All the ``pictures'' I can come up with [via purely formal reasoning] sit in at least
% 5 dimensions -- one for each of the wires, and another to give room to maneuver.}
% \end{comment}

@@

There are two different interpretations for an operational semantics
for the language of equivalences:
\begin{enumerate}
\item Mimicking closely the one in the previous section, and
thus finding explicit homotopies between the functions induced
by the operational semantics of the level-1 combinators.
\item Treating things more syntactically, and interpreting
the combinators as program transformations.
\end{enumerate}
A previous paper~\cite{Carette2016} explores the first interpretation
in depth. There one can find a definition of ``equivalences of
equivalences'', which as the base of that interpretation.

@@

>>>>

\documentclass{article}
\usepackage[utf8x]{inputenc}
\usepackage{agda}
\usepackage[nocenter]{qtree}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{fullpage}
\usepackage{url}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{listings}
\usepackage{multicol}
\usepackage{stmaryrd}
\usepackage{proof}
\usetikzlibrary{quotes,fit,positioning}
\tikzset{>=latex}

@@

\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
\newcommand{\alt}{~|~}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\ensuremath{\mathit{unite}_+\mathit{l}}}
\newcommand{\identrp}{\ensuremath{\mathit{uniti}_+\mathit{l}}}
\newcommand{\identlsp}{\ensuremath{\mathit{unite}_+\mathit{r}}}
\newcommand{\identrsp}{\ensuremath{\mathit{uniti}_+\mathit{r}}}
\newcommand{\swapp}{\ensuremath{\mathit{swap}_+}}
\newcommand{\assoclp}{\ensuremath{\mathit{assocl}_+}}
\newcommand{\assocrp}{\ensuremath{\mathit{assocr}_+}}
\newcommand{\identlt}{\ensuremath{\mathit{unite}_*\mathit{l}}}
\newcommand{\identrt}{\ensuremath{\mathit{uniti}_*\mathit{l}}}
\newcommand{\identlst}{\ensuremath{\mathit{unite}_*\mathit{r}}}
\newcommand{\identrst}{\ensuremath{\mathit{uniti}_*\mathit{r}}}
\newcommand{\swapt}{\ensuremath{\mathit{swap}_*}}
\newcommand{\assoclt}{\ensuremath{\mathit{assocl}_*}}
\newcommand{\assocrt}{\ensuremath{\mathit{assocr}_*}}
\newcommand{\absorbr}{\ensuremath{\mathit{absorbr}}}
\newcommand{\absorbl}{\ensuremath{\mathit{absorbl}}}
\newcommand{\factorzr}{\ensuremath{\mathit{factorzr}}}
\newcommand{\factorzl}{\ensuremath{\mathit{factorzl}}}
\newcommand{\dist}{\ensuremath{\mathit{dist}}}
\newcommand{\factor}{\ensuremath{\mathit{factor}}}
\newcommand{\distl}{\ensuremath{\mathit{distl}}}
\newcommand{\factorl}{\ensuremath{\mathit{factorl}}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\sem}[1]{\ensuremath{\llbracket{#1}\rrbracket}}

@@

\end{abstract}

@@

@@

@@

@@

\begin{center}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
  \draw[fill] (-2,-1.5) circle [radius=0.025];
  \node[below] at (-2.1,-1.5) {$A$};
  \node[below] at (-2.1,-1.9) {$+$};
  \draw[fill] (-2,-2.5) circle [radius=0.025];
  \node[below] at (-2.1,-2.5) {$B$};

  \draw[fill] (6.5,-1.5) circle [radius=0.025];
  \node[below] at (6.7,-1.5) {$C$};
  \node[below] at (6.7,-1.9) {$+$};
  \draw[fill] (6.5,-2.5) circle [radius=0.025];
  \node[below] at (6.7,-2.5) {$D$};

@@

  \path (1.5,0.5) node (tc1) [func] {$c_1$};
  \path (1.5,-0.5) node (tc2) [func] {$c_2$};
  \path (3,-4.5) node (bc1) [func] {$c_1$};
  \path (3,-3.5) node (bc2) [func] {$c_2$};
\end{tikzpicture}
\end{center}
The top path is the $\Pi$ program
$(c_1~\oplus~c_2)~\odot~\swapp$ which deforms the
type $A$ by $c_1$, deforms the type $B$ by $c_2$, and deforms the
resulting space by a twist that exchanges the two injections into the
sum type. The bottom path performs the twist first and then deforms
the type $A$ by $c_1$ and the type $B$ by $c_2$ as before. One
could imagine the paths are physical \emph{elastic} wires in $3$ space, where
the deformations $c_1$ and $c_2$ as arbitrary deformations on these wires, and
the twists do not touch but are in fact well-separated. Then, holding the
points $A$, $B$, $C$, and $D$ fixed, it is possible to imagine
sliding $c_1$ and $c_2$ from the top wire rightward past the
twist, and then using the elasticity of the wires, pull the
twist back to line up with that of the bottom --- thus making
both parts of the diagram identical.  Each of these moves
can be undone (reversed), and doing so would take the bottom
part of the diagram into the top part.  In other
words, there exists a deformation of the program
$(c_1~\oplus~c_2)~\odot~\swapp$ to the program
$\swapp \odot (c_2~\oplus~c_1)$. We can also show that this
means that, as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and
$\swapp \odot (c_2~\oplus~c_1)$ are equal. And, of course, not
all programs between the same types can be deformed into one
another. The simplest example of inequivalent deformations
are the two automorphisms of $1+1$, namely $\idc$ and $\swapp$.

@@

Previously we saw how we could take the proof terms of commutative semiring
equivalences as our starting point for $\Pi$. What we need
now is to understand how \emph{proofs} of algebraic identities should be
considered equivalent. Classical algebra does not help, as proofs
are not considered first-class citizens. However,
another route is available to us: since the work of
Hofmann and Streicher~\cite{hofmann96thegroupoid}, we know that
one can model types as \emph{groupoids}.  The additional
structure comes from explicitly modeling the ``identity
types'': instead of regarding all terms which witness
the equality of (say) $a$ and $b$ of type $A$ as being
indistinguishable, we posit that there may in fact be many.
This consequences of this one decision are enough to show that
types can be modeled by groupoids.

Thus, rather than looking at (untyped) commutative semirings, we
should look at a \emph{typed} version. This process frequently goes by
the moniker of ``categorification.''  We want a categorical algebra,
where the basic objects are groupoids (to model our types), and where
there is a natural notion of $+$ and $*$.  At first, we hit what seems
like a serious stumbling block: the category of all groupoids, \Gpd,
have neither co-products nor products. However, we don't want to work
internally in \Gpd -- we want operations \emph{on} groupoids. In other
words, we want something akin to symmetric monoidal categories, but
with two interacting monoidal structures.  Luckily, this already
exists: the categorical analog to (commutative) semirings are
(symmetric) Rig Categories~\cite{laplaza72,kelly74}.  This
straightforwardly generalizes to symmetric Rig Groupoids.

How does this help? Coherence conditions! Symmetric monoidal categories,
to start somewhere simple, do not just introduce natural transformations
like the associator $\alpha$ and the left and right unitors ($\lambda$
and~$\rho$ respectively), but also coherence conditions that these must satisfy.
Looking, for example, at just the additive fragment of $\Pi$ (i.e. with just $0$,
$1$ and $+$ for the types, $\odot$ and $\oplus$ as combinators, and
only the terms so expressible), the sub-language would correspond,
denotationally, to exactly (non-empty) symmetric monoidal groupoids. And what
these possess are exactly some \emph{equations between equations}
as commutative diagrams.  Transporting these coherence conditions, for
example those that express that various transformations are \emph{natural}
to $\Pi$, gives a list of equations between $\Pi$ programs.
Furthermore, all the natural transformations
that arise are in fact natural \emph{isomorphisms} -- and thus
reversible.

@@

Starting with the simplest constructions, the first two constructs in
Fig.~\ref{fige} are the level-2 analogs of~$+$ and~$*$, which
respectively model level-1 choice composition and parallel composition
(of equivalences).  These allow us to ``build up'' larger equivalences
from smaller ones.  The next two express that both of these
composition operators distribute over sequential composition $\odot$
(and vice versa).

@@

The bottom line in Fig.~\ref{figj} is actually a linear
restatement of the famous ``pentagon diagram'' stating a
particular coherence condition for monoidal categories~\cite{KELLY197197}.
To make the relation between $\Pi$ as a language and the
language of category theory, the figure below displays
the same morphism but in categorical terms.

@@

@@

There are two different interpretations for an operational semantics
for the language of equivalences:
\begin{enumerate}
\item Mimicking closely the one in Sec.~\ref{sec:opsem}, and thus
  finding explicit homotopies between the functions induced by the
  operational semantics of the level-1 combinators.
\item Treating things more syntactically, and interpreting
the combinators as program transformations.
\end{enumerate}
A previous paper~\cite{Carette2016} explores the first interpretation
in depth. There one can find a definition of ``equivalences of
equivalences'', which as the base of that interpretation.

@@

\medskip
{\footnotesize 
